---
title: "R and Time series"
output: html_notebook
---

R is a free, open-source programming language widely used for statistical analysis and data visualization. It is best used with an Integrated Development Environment (IDE) like RStudio.

To get started with R, you must install the R programming language first, followed by the RStudio interface.
https://posit.co/download/rstudio-desktop/


## Core Concepts 

**Objects and Assignment**: In R, you store values in objects using the assignment operator `<-`. 

```{r}
# Assign the value 10 to an object named 'x'
x <- 10
# Print the value of x
print(x)

y <- 100
y
```
**Functions**: Functions are blocks of reusable code that perform specific tasks. To use a function, you pass arguments within parentheses. 

```{r}
# Use the sum() function to add numbers in a vector
sum(c(1, 2, 3))
```

**Packages**: Packages are collections of functions and data that extend R's capabilities. You must install a package once and load it into your session with `library()` every time you use it.  

```{r, eval=FALSE}
# Install a package (only needs to be done once)
# install.packages("ggplot2")
# Load the package into the current session
library(ggplot2)
```

## Data Types and Structures 

R uses various data types and structures: 

### Data Types: 

**Numeric**: For numbers (e.g., 10 , 3.14 ). 

**Character**: For text (e.g., "hello" ). 

**Logical**: For `True` or `FALSE` values. 

**Factor**: For categorical data. 

### Data Structures: 

**Vectors**: One-dimensional structures that store elements of the same data type. You create them using the `c()` function.
	
```{r}
	# A numeric vector
my_vector <- c(1, 2, 3, 4)
```
	
**Matrices**: Two-dimensional arrays that store elements of the same data type. 

**Lists**: One-dimensional structures that can store elements of different data types or even other data structures. 

**Data Frames**: The most commonly used structure for data analysis; they are two-dimensional, like a spreadsheet, where each column can have a different data type but all columns must have the same length. You can create a data frame using the `data.frame()` function or import data from a file using functions like `read.csv()`.   

### Basic Operations 

• Arithmetic: Standard operators like `+`,`-`,`*` ,`/` ,`^`  (exponent) are available. 

• Data Manipulation: Use functions like `subset()` to filter data,  `head()` to view the first few rows, and `summary()` to get statistical summaries. 

• Getting Help: To get help on any function, type `?` followed by the function name in the console (e.g., `?subset` ).   



## Import Time Series Data

Data in the file "s&p_2026.csv" is the weekly S&P 500 index from Jan 24, 2025- Jan 21, 2026, obtained from https://www.marketwatch.com/investing/index/spx/download-data.

Let's first import the dataset in R.

#### Step 1: Set working directory where your data is saved.

#### stpe 2: Import dataset with the following code. 
```{r}
snp<-read.csv("s&p_2026.csv" )
```

To check the type of data:

```{r}
class(snp)
```
To check how the data looks like:


```{r}
# to see the first few rows
head(snp)
```


```{r}
# to see the last few rows
tail(snp)
```


To choose a specific column in the dataset 

```{r}
open_price<-snp$Open
```

```{r}
class(open_price)
head(open_price)
```
To visualize the open price of S\&p500 over the time of the dataset

```{r}
plot(open_price, type="l")
```

This is called time series plot.

## Set as time series object and draw time series plot

R package `fpp2`: provides various datasets (e.g., `a10` for drug sales, `ausbeer` for beer production, `goog` for stock prices) and tools useful in time series analysis. 

```{r, eval=FALSE}
library(fpp2)
```

Make a numeric vector into time series object using a R-function `ts()`

```{r}
SP_ts<- ts(open_price ,start=c(1,3), freq = 4)
#or
SP_ts2<-ts(open_price ,start=1, freq = 4)
```


`freq`=frequency, number of observations in each cycle/season

`start`=	the time of the first observation; should be a vector of two numbers, or just one number. 

To know more about the function, try
```{r}
?ts
```


```{r}
class(SP_ts)
```


```{r}
head(SP_ts)
tail(SP_ts)
```
```{r}
head(SP_ts2)
```

```{r}
autoplot(SP_ts, ylab="Stock price")
```

```{r}
autoplot(SP_ts2)
```

Or convert the whole dataframe into time series object:

```{r}
SP_ts3<- ts(snp ,start=1, freq = 4)
class(SP_ts3)
```
```{r}
head(SP_ts3)
```


```{r}
SP_ts3[,1]
SP_ts3<-SP_ts3[,-1]
```
```{r}
head(SP_ts3)
```
```{r}
autoplot(SP_ts3)
```


## Obtain stock prices in R

To obtain stock prices in R, use the `quantmod` package for historical data with `getSymbols()`


```{r, warning=FALSE}

install.packages("quantmod")
library(quantmod)

# Get Apple stock data
getSymbols("AAPL", from = "2023-01-01", to = "2024-01-01")

# View the first few rows
head(AAPL)

# Plot the stock price
chartSeries(AAPL)
```


```{r}
getSymbols("AAPL", 
           src = "yahoo", 
           from = "2025-01-01", 
           to = "2026-02-24", 
           periodicity = "weekly")

chartSeries(AAPL)
```

Periodicity Argument: Supports "daily", "weekly", and "monthly". It is efficient because it downloads a smaller volume of data from the web.

```{r}
# loads Ford as time series class ts 
getSymbols('F',src='yahoo',from = "2025-01-01", 
           to = "2026-02-24", 
           periodicity = "daily",
           return.class='ts') 
chartSeries(F)
autoplot(F)
autoplot(F, facets = TRUE)
```


## Types of variations in time series data

types of variations

**Trend**: general changes observed in mean-level.

**Cyclical or seasonal variation**: repeated patterns observed with cycles or seasons (e.g. cyclical patterns in monthly sales of electronics, monthly average temperature, monthly sales of heater, monthly electric bill.)

**Other irregular fluctuations**: If it is purely random fluctuation, it is not of our interest—we cannot analyze it further. If it has a structure in the fluctuations—we will find the structure and model it with time series techniques.


# Exercise
1. Extract the monthly retail trade and food services data –not seasonally adjusted---from 1992 to 2026 from the website below.
https://www.census.gov/econ/currentdata/
Save the file in csv form (you need to clean up the dataset in Microsoft Excel) and read it in R studio. Draw time series plot with the data. Attach the plot. Describe types of fluctuations that are observed from the data.
Is there any trend observed over time?
Any seasonal variations?
Irregular fluctuation?

Hint:

```{r}
data_retail<-read.csv("retail.csv")

head(data_retail)
tail(data_retail)
class(data_retail)
```
Our variable `Value` should be numeric, however...

```{r}
class(data_retail$Value)
```

How to change the character vector into numeric vector?

method 1: change it to numbers in Microsoft Excel and import the data again.

method 2: covert to numeric in R as follows

```{r}
# Your character vector
my_vector <- c("10.5", "20", "30.7")
class(my_vector)
# Convert to numeric
my_numeric_vector <- as.numeric(my_vector)
class(my_numeric_vector)
```
If we try
```{r}
as.numeric(data_retail$Value)
```


It does not work, since the characters include `,`.


Clean-up: If your vector has commas (e.g., "1,200"), use gsub() before converting:

For example,
```{r}
as.numeric(gsub(",", "", "1,200"))
```
For our case:
```{r}
ourdata<-as.numeric(gsub(",", "", data_retail$Value))
head(ourdata)
```
Now we are ready to make it into time series object:
since it is monthly data which starts from January 1992,

```{r}
ts(ourdata, start=c(1992, 1), frequency = 12)
```










