---
title: "Time series (and data frames)"
output: html_notebook
---

```{r}
library(fpp2)
```

Time series are everywhere, any signal indexed by time is a time series. Examples:

* Call volumes to a customer service center
* stock price
* housing price
* temperature throughout a day
* ...

The goal of time series analysis is to forecast!


# 1. Define a single variable time series
`ts(data =, start =, frequency =)`
```{r}
observations <- c(100, 20, 30, 55, 45, 70, 90, 24, 52, 35, 60, 73, 18, 25)
y_month <- ts(observations, start = c(2000, 2), frequency = 12)
y_month
```
- frequency = 5 or 7: daily data
- frequency = 4: quarterly data
- frequency = 12: monthly data
- frequency = 52: weekly data

```{r}
y_quart <- ts(observations, start = c(1990,3), frequency=4)
y_quart
```

```{r}
y_daily <- ts(observations, start=c(1,3) ,frequency=5)
y_daily
```
```{r}

y_year <- ts(observations, start = 2000)
y_year
```

to know more about `ts`, use r-code `?ts` or `help(ts)`.


# 2. Plot
The `autoplot()` function in package `ggplot2` is a convenient way to generate various types of plots based on the type of data object passed to it.
```{r}
autoplot(y_month)
autoplot(y_quart)
autoplot(y_daily)
autoplot(y_year)
```


## add x/y labels and title
```{r}
autoplot(y_year) +
  labs(title = "A title",
       x = "Year",
       y = "Value")
```


## Explore other time series that come with fpp2 {#explore}

type `help(fpp2)` in console and then click "Index" at the bottom.
```{r}
# plot a10
autoplot(a10) +
  labs(title = "Monthly anti-diabetic drug subsidy in Australia from 1991 to 2008")
```

# 3. Useful methods to explore time series
- `start()`, `end()`, `frequency()`
- `window()`: very useful in truncating a time series because it retains the time series attributes.

explore this a10 time series object a bit more.

```{r}
end(a10)
frequency(a10)
```

```{r}
window(a10, start = c(2008, 5))
```

*Below is a bad way to truncate a time series, unless you only want the raw values.*
```{r}
n <- length(a10)
a10[(n - 1) : n]
```

# 4. Useful methods to explore dataframes and time series
- `head()`: view the first few rows (or entries)
- `tail()`: similar but for the end

```{r}
head(a10, 3)
```

## for data frames
```{r}
df <- data.frame(name = c("A", "B", "C"),
                 height = c(160, 170, 165))
df
```

```{r}
head(df, 2)
```

Retrieve entries in a data frame

```{r}
# first row, 2nd col
df[1, 2]
```

```{r}
df[2:3, 2]
```

```{r}
df[c(1, 3), 'name']
```
```{r}
df[2,]
df[,2]
```

## elecdemand: a multi-variable time series
`View(elecdemand)` will open a separate tab for us to view this data set.

```{r}
help(elecdemand)
autoplot(elecdemand)
```

retrieve entries from elecdemand


get a specific column: `df[, col_idx]`
```{r, eval=FALSE}
elecdemand[, 2]
```

get only first (Demand) and third (Temperature) column of `elecdemand`
```{r eval=FALSE}
 elecdemand[, c(1, 3)]
 elecdemand[, c('Demand', 'Temperature')]
```


get first row
```{r, eval=FALSE}
 elecdemand[1, ]
```

### draw a scatter plot for a multi-variable time series
`ggplot()` needs a data frame, but elecdemand is not a data frame object.
```{r}
class(elecdemand)
```
msts is an object for multi seasonal time series objects, intended to be used for models that support multiple seasonal periods

```{r}

head(elecdemand)
```


```{r}
ed_df <- as.data.frame(elecdemand)
class(ed_df)
```

scatter plot
```{r}
ggplot(data = ed_df,
       mapping = aes(x = Temperature, y = Demand)) +
  geom_point()
```

more exercise on plot: drawing only first three rows

the later two functions inherit the data and aes in `ggplot()`.
```{r}
ggplot(data = ed_df[1:3,],
       mapping = aes(x = Temperature, y = Demand)) +
  geom_point() +
  geom_line()
```


```{r}
ggplot(data = ed_df[1:3,],
       mapping = aes(x = Temperature, y = Demand)) +
  geom_point(data = ed_df[4:6,]) +
  geom_line()
```

# 5. Time series patterns
* Trend: A trend exists when there is a long-term increase or decrease in the data. It does not have to be linear.

  In the `a10` [plot](#explore), there is an upward trend.
  

* Seasonal: Seasonality is always of a **fixed** and known frequency.

  In the `a10` plot, there is also a seasonal pattern.

* cyclic: A cycle occurs when the data exhibit rises and falls that are **not of a fixed** frequency. It is like seasonal, but each cycle's length will vary.

## Explore seasonality with seasonal plots
- `ggseasonplot()` 


```{r}
ggseasonplot(a10)
```

Recall `window()` subsets a time series nicely.
```{r}
ggseasonplot(window(a10, end = c(1993, 12)))
```



## A cyclic example
```{r}
autoplot(lynx) 
frequency(lynx)
head(lynx)
```

The following code will find all the local max (peaks) of a time series.

```{r}
# You can explore this code on your own. It will not be tested.
which(diff(sign(diff(lynx))) == -2) + 1
```
















