---
title: "Time Series Linear Regression"
output: html_notebook
---



# 1. Use other features as predictors


# $$ y_t=x_{1t}+x_{2t}+\cdots +x_{kt}+e_{t}  $$

`tslm()` is similar to `lm()`, but has extra functionality for time series.

plot `uschange`
```{r}
library(fpp2)
autoplot(uschange, facets = TRUE)
```


We can think of a (multi-variate) time series as a data frame **where each row is ordered**.


Consumption ~ Income + Savings + Unemployment


```{r}
fitISU <- tslm(Consumption ~ Income + Savings + Unemployment, data = uschange)
summary(fitISU)
```

Fitted model equation:

$$\hat y_t= 0.28+ 0.73x_{1t}-0.05 x_{2t} -0.34 x_{3t},$$ where

$y_t$=Consumption at time $t$

$x_{1t}$=Income at time $t$

$x_{2t}$=Savings at time $t$

$x_{3t}$=Unemployment at time $t$



Model diagnostics with residuals:
```{r}
hist(fitISU$residuals)
hist(rstudent(fitISU))
plot(y=rstudent(fitISU),x=as.vector(time(uschange)),
xlab='Time',ylab='Standardized Residuals',type='o')
qqnorm(rstudent(fitISU))
plot(y=rstudent(fitISU),x=as.vector(fitted(fitISU)),
xlab='Fitted Trend Values',
ylab='Standardized Residuals')
```

The `forecast` function is similar to `predict()`.


For the future values of Income, Savings and Unemployment, each would need to have its own forecast. Let's suppose that we have already forecasted the future 2 values of these 3 variables.

```{r}
futureISU <- data.frame(Income = c(1, 1.5),
                      Savings = c(5, 5.5),
                      Unemployment = c(0.1, -0.1))
forecast1 <- forecast(fitISU, newdata = futureISU, level = 0.95)
forecast1
```

plot Consumption and the forecast
```{r}
autoplot(uschange[, 'Consumption']) +
  autolayer(forecast1, PI = FALSE)
```

# 2. Create time-dependent predictors 
Let's use 
```{r}
autoplot(euretail)
```


## linear trend
model $y_t=\beta_0+\beta_1*t+\epsilon_t$
```{r}
N <- length(euretail)
t <- 1:N
fit.t <- tslm(euretail ~ t)
```

Plot fitted values
```{r}
autoplot(euretail) +
  autolayer(fitted(fit.t), series = "fitted value")
```

## linear trend: R
`tslm(y ~ trend)`
```{r}
fit.t2 <- tslm(euretail ~ trend)
cbind(fitted(fit.t), fitted(fit.t2))[1:4,]
```


## add seasonality 
need to create dummy variables (or indicator variables) for each quarter

q1 = 1 if it is quarter 1, = 0 elsewhere
```{r}
head(euretail) # y started at Q1
```

```{r}
q1 <- rep(c(1, 0, 0, 0), length.out = N)
```


create q2, q3 similarly
```{r}
q2 <- rep(c(0, 1, 0, 0), length.out = N)
q3 <- rep(c(0, 0, 1, 0), length.out = N)
```


q4 is represented by all 0's. we do not need to create q4 because we have an intercept.

```{r}
# use both trend and seasonality as predictors
fit.t.s <- tslm(euretail ~ t + q1 + q2 + q3)
autoplot(euretail) +
  autolayer(fitted(fit.t.s))
```

forecast using the model `fit.t.s`
```{r}
h <- 4
forecast(fit.t.s, newdata = data.frame(t = seq(N + 1, N + h),
                                      q1, q2, q3))

```


## add seasonality: R
`tslm(y ~ trend + season)`
```{r}
fit.t.s2 <- tslm(euretail ~ trend + season)
summary(fit.t.s2)
```

forecast is easier too if using built-in R function.
```{r}
forecast(fit.t.s2, h = 4)
```



## More use of indicator variables
* model interventions such as holidays, competitor activity, industrial action, etc. Add a variable that equals 1 on the special event and 0 elsewhere. If the intervention causes a level shift, then we will let it equal 1 on the event and onward.
* deal with outliers. Add a variable that equals 1 at the outlier location and 0 elsewhere.



## Piecewise linear fit
When we see clear historical break points in a time series, we can fit it with piecewise linear functions.

These points, where the slopes are changing, are called knots. Suppose the knots are $c_2, c_3, \cdots, c_M$, then we are going to define M regressors (predictors) as follows
$$x_1 = t, x_2=(t-c_2)_+, x_3 = (t-c_3)_+, \cdots, x_M = (t-c_M)_+$$
The notation $(t-c)_+$ means it equals 0 for $t<c$ and equals $t-c$ elsewhere.

```{r}
autoplot(euretail)
```

```{r}
# find c2
c2 <- which.max(euretail)
c2
```

pmax will return point-wise max (compared two vectors coordinate-wise)
```{r}
pmax(c(1, 2, -1), c(0, 0, 0))
```

At index 48, the vector is obtaining its maximum.
```{r}
# make x1, x2
x1 <- t
x2 <- pmax(t - c2, 0)
```

fit with x1, x2
```{r}
fit.pl <- tslm(euretail ~ x1 + x2)
autoplot(euretail)+
  autolayer(fitted(fit.pl), series = "Fitted value")
```

```{r}
summary(fit.pl)
```

# Exercise
1. Use "uschange" dataset with response, Consumption, and predictors, Income, Savings, Unemployment, trend and season, to fit a linear regression. Check the model fit with residuals, and compare how the model has changed (estimates and significance of the parameters, AIC, BIC) from what you obtained in 1 in this notebook. 
2. Use "euretail" dataset and fit a linear regression with predictors, piecewise linear trend and season. Make forecasts for the next 10 observations.  Draw a plot with original response overlaid with the predicted values. 
